---
title: 'Assignment on Data Science for Good: Center for Policing Equity'
author: 'Name: YASWANTH MARAM, Mail: ym20741@essex.ac.uk'
date: "February 19th, 2021"
output:
  word_document: default
  html_document: default
---

### Introduction:

We are given data which has Crime Incident cases happened in Dallas, USA. The data consist of 2384 incidences with available information on 47 variables providing us information on the incidents, dates, officers, subjects, locations, race, time, and other metrics. We will investigate the data by providing descriptive statistics for all relevant variables and formulate some interesting questions and try to answer them by doing exploratory data analysis. 

```{r setup, include=FALSE}
library(ggmap)
library(shiny)
library(leaflet)
library(RColorBrewer)
library(tidyverse)
library(leaflet.extras)
library(lubridate)
library(gridExtra)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggridges)
library(viridis)
#knitr::opts_chunk$set(echo = TRUE)
```

#### Interesting Questions:

1.	How is the spread of incidents across year, months, hours? Is there any trend?
2.	Study of incident/crime rates across different subject Races?
3.	Is there any disparity between races in Officers and Subjects?
4.	Is the crime rate increasing or decreasing among division?
5.	Are there any specific reasons for arrests?
6.	Is it more likely for an officer with a few years in the force to use violence?

```{r echo=FALSE}
# Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
df <- read.csv("C:\\Users\\yaswa\\Desktop\\EDA\\Assignment1 Yash\\37-00049_UOF-P_2016_prepped.csv",na.strings = c(""))
df = df[-1,]
```

### Exploratory Analysis:

#### 1.Spread of incidents across year, months, hours:

```{r echo=FALSE, message=FALSE, warning=FALSE,results='hide'}
df$INCIDENT_DATE <- as.Date(df$INCIDENT_DATE, format = "%m/%d/%Y")
df$INCIDENT_DATE <- gsub("00","20",df$INCIDENT_DATE)
df$INCIDENT_DATE <- as.Date(df$INCIDENT_DATE, format = "%Y-%m-%d")
df$INCIDENT_TIME <- format(strptime(df$INCIDENT_TIME, "%I:%M:%S %p"), "%H:%M:%S")
df$INCIDENT_MONTH <- months(as.Date(df$INCIDENT_DATE))
df$INC_MONTH <-format(df$INCIDENT_DATE,"%m")
df$INCIDENT_HOUR <- as.numeric(substr(df$INCIDENT_TIME, 0, 2))
df$INCIDENT_DAY <- wday(df$INCIDENT_DATE, label=TRUE)
df$INC_HOUR <- substr(df$INCIDENT_TIME, 0, 2)
df$INC_DATE <- substr(df$INCIDENT_DATE, 9, 10)
## Create group of datas:

df_year <-  df %>%
  group_by(INCIDENT_DATE,INCIDENT_MONTH,INCIDENT_DAY) %>%
  summarize(count = n())

df_month <-  df %>%
  group_by(INC_MONTH) %>%
  summarize(count = n())

df_day <-  df %>%
  group_by(INCIDENT_DAY,INCIDENT_HOUR) %>%
  summarize(count = n())

df$INC_HOUR <- substr(df$INCIDENT_TIME, 0, 2)

df   %>% group_by(INC_HOUR) %>%
  summarize(avg =n()) -> df_hour_n
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
c1 <- ggplot(data = df_year, aes(INCIDENT_DATE, count)) +   geom_line(size=0.5, col="gray") +
geom_smooth(method = "loess", color = "red", span = 1/5) + theme_bw() + labs(x="Months ", y= "INCIDENT COUNTS", title="1.a Year vs Incidents")

r1 <- ggplot(df_month, aes(x=INC_MONTH, y =count, group=1)) + geom_line()  + geom_line( size = 1,colour ="steelblue") + labs(x="MONTHS OF 2016", y= "INCIDENT COUNTS", title="1.b Months vs  Incident Rates")  + theme_bw()


r2 <- ggplot(df_hour_n, aes(x = INC_HOUR, y = avg, group = "count")) + geom_line( size = 1, colour = "orange") + labs(x="HOURS IN A DAY", y= "INCIDENT COUNTS", title="1.c Hours vs  Incident Rates")+ theme_bw() +
theme(axis.text.x=element_text(angle=-90, vjust=0.5)) +

  labs(x = "Hour of the day", y = "count") + theme_bw()

r3 <- ggplot(df_year, aes(count)) +
  geom_density(alpha = 0.5, colour = "black", fill ="blue")+ labs(x="Incident counts", y= "Density", title="1.d Distribuion of incident rates") + theme_bw()


grid.arrange(c1,r1,nrow=2)
#c1
#r1
#r2
#r3
```

*	From figure 1.a and 1.b it is very clear that the incidents have been gradually reducing in the year 2016.
*	March has the highest number of incidents, while December has the least number. This may be the result of due to seasonality (holidays).

```{r echo=FALSE}

r2 <- ggplot(df_hour_n, aes(x = INC_HOUR, y = avg, group = "count")) + geom_line( size = 1, colour = "orange") + labs(x="HOURS IN A DAY", y= "INCIDENT COUNTS", title="1.c Hours vs  Incident Rates")+ theme_bw() +
theme(axis.text.x=element_text(angle=-90, vjust=0.5)) +

  labs(x = "Hour of the day", y = "count") + theme_bw()
r2
```

*	From figure 1.c we can clearly see that the number of incidents reported were lowest at 6a.m & 7a.m and highest at 5p.m & 8p.m

```{r echo=FALSE}
r3 <- ggplot(df_year, aes(count)) +
  geom_density(alpha = 0.5, colour = "black", fill ="blue")+ labs(x="Incident counts", y= "Density", title="1.d Distribuion of incident rates") + theme_bw()
r3
```

*	From figure 1.d we can clearly see right skewness in the incident count. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggridges)

###Multiplot function to join graphs:

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}




###Back to analysis:

p1 <- ggplot(df_year, aes(count, INCIDENT_MONTH, fill = INCIDENT_MONTH)) +
  
  geom_density_ridges(bandwidth = 0.1,size = 1) +
#  scale_fill_viridis(name = "T_max [°C]", option = "C") +
  scale_x_log10() +
  
  theme(legend.position = "none") +
  
  labs(x = "Number of Incidents", y = "Months")

#p1


p2 <- ggplot(df_year, aes(count, INCIDENT_DAY, fill = INCIDENT_DAY)) +
  
  geom_density_ridges(bandwidth = 0.1,size = 1) +
  
  scale_x_log10() +
  
  theme(legend.position = "none") +
  
  labs(x = "Number of Incidents", y = "Days")

#p2


layout <- matrix(c(1,1,2,2,2),1,5,byrow=TRUE)

multiplot(p1, p2, layout=layout)
```

* The above figures show the number of Incidences plotted against Months and days. We can see skewness in the months of September, December, February, May and October.
* The mean range of incidents is between 3 to 10 for both days and months.


#### 2.	Study of Incident rates for Subject Races:

Boxplot analysis will give us more intuition about the quartile ranges, medians and outliers. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
df   %>% 
  filter(SUBJECT_RACE == "Black"  | SUBJECT_RACE == "White" | SUBJECT_RACE == "Hispanic" ) %>%
  group_by(INCIDENT_DATE,INC_MONTH,SUBJECT_RACE) %>%
  summarize(avg =n()) -> df_dateh


g3 <- ggplot(df_dateh , aes(x = (INC_MONTH), y= avg, fill = INC_MONTH)) +
  
      geom_boxplot() +
      labs(x= 'days',y = 'Incident Rate', 
           title = paste("Central Tendency of", ' Incident rate across SUBJECT RACE ')) +
      # theme_bw() +
  theme(legend.position="none") + facet_wrap(~SUBJECT_RACE)  + 
    coord_cartesian(ylim = c(1, 12))
g3
```

* From the graphs above it is evident that the black people have more incident rates compared to Hispanic and whites. 
* More outliers are observed with Hispanic and white population than blacks.

#### 3. Officers vs Subjects Races Analysis:

```{r echo=FALSE, message=FALSE, warning=FALSE}
df_srace <-  df %>%
  group_by(INC_MONTH,INCIDENT_MONTH,SUBJECT_RACE) %>%
  summarize(count = n())

df_orace <-  df %>%
  group_by(INC_MONTH,INCIDENT_MONTH,OFFICER_RACE) %>%
  summarize(count = n())



c1 <- ggplot() +
geom_line(data=subset(df_srace, SUBJECT_RACE=="Black" ) ,aes(y=count,x= INC_MONTH,colour="green",group=1),size=1 ) +
geom_line(data=subset(df_srace, SUBJECT_RACE=="Hispanic" ) ,aes(y=count,x= INC_MONTH,colour="red",group=1),size=1 ) +
  geom_line(data=subset(df_srace, SUBJECT_RACE=="White" ) ,aes(y=count,x= INC_MONTH,colour="orange",group=1),size=1 ) +
  scale_color_discrete(name = "Y series", labels = c("BLACKS", "WHITES","HISPANICS")) + labs(x="Months of 2016", y= "Counts", title=" Subject Race vs Incidents")  +
  scale_color_discrete(name = "Legend", labels = c("BLACKS", "WHITES","HISPANICS"))  + theme(axis.text.x=element_text( vjust=0.5),legend.position="bottom") +guides(colour=guide_legend(nrow=2))

c2 <- ggplot() +
  geom_line(data=subset(df_orace, OFFICER_RACE=="Black" ) ,aes(y=count,x= INC_MONTH,colour="green",group=1),size=1 ) +
  geom_line(data=subset(df_orace, OFFICER_RACE=="Hispanic" ) ,aes(y=count,x= INC_MONTH,colour="red",group=1),size=1 ) +
  geom_line(data=subset(df_orace, OFFICER_RACE=="White" ) ,aes(y=count,x= INC_MONTH,colour="orange",group=1),size=1 ) +
  scale_color_discrete(name = "Y series", labels = c("BLACKS", "WHITES","HISPANICS")) + labs(x="Months of 2016", title=" Officer Race vs Incident Handled Rates")  +
  scale_color_discrete(name = "Legend", labels = c("BLACKS", "WHITES","HISPANICS"))  + theme(axis.text.x=element_text( vjust=0.5),legend.position="bottom")+guides(colour=guide_legend(nrow=2))


layout <- matrix(c(1,1,1,1,2,2,2,2),3,8,byrow=TRUE)

multiplot(c1, c2, layout=layout)

```

In the above graph we can clearly see that the number of black and Hispanic officers handling the incidents is very less compared to white officers while for subjects it is quite opposite.

#### 4.	Division analysis:

In the below figure the 7 divisions have been plotted against the months in x-axis and incident counts in y-axis.

```{r echo=FALSE, message=FALSE, warning=FALSE}
df_div <-  df %>%
  group_by(INC_MONTH,DIVISION) %>%
  summarize(count = n())


#head(df_div,20)

library(scales)

ggplot(df_div, aes(x = INC_MONTH, y = count, group = 'count'))  + geom_line( aes(color= DIVISION), size = 1.2) + #labs(x="HOURS IN A DAY", y= "INCIDENT COUNTS", title=" Hours vs No of Incidents") 
   facet_wrap(~ DIVISION,ncol=4) + #+ theme(axis.text.x=element_text(angle=-90, vjust=0.5)) +
theme(legend.position="none") + 
  labs(x="Months ", y= "INCIDENT COUNTS", title=" Division vs Incidents")
```

We can clearly see that the central division has the greatest number of crimes reported. A neutral trend has been observed from northeast and north central divisions. Steep slopes are observed for northwest, south central, southeast, and southwest divisions.

#### 5.	Reasons for Arrest distributions:

In the below graph we can observe the distributions of rates of various subject descriptions i.e. reasons for arrests across the year.

```{r echo=FALSE, message=FALSE, warning=FALSE}
df_subdes <-  df %>%
  group_by(INCIDENT_DATE,INCIDENT_MONTH,INCIDENT_DAY,SUBJECT_DESCRIPTION) %>%
  summarize(count = n())


library(viridis)
ggplot(subset(df_subdes,SUBJECT_DESCRIPTION != "FD-Motor Vehicle" & SUBJECT_DESCRIPTION != "NULL" & SUBJECT_DESCRIPTION != "FD-Animal" & SUBJECT_DESCRIPTION != "Animal"), aes(x = INCIDENT_DATE, y = SUBJECT_DESCRIPTION, fill = ..x..)) +
  
  geom_density_ridges_gradient(size=0.7,scale = 1, rel_min_height = 0.01, gradient_lwd = 1., bandwidth = 6,alpha = 1.5) +
   scale_fill_viridis(name = "Tail probability", direction = -1) +
  
  ggtitle("Distribution of SUBJECT DESCRIPTION") +
  
  labs(x = "months of 2016", y = "", fill = "Humidity") +
  
  theme_ridges(font_size = 13, grid = TRUE) +
  
  theme(legend.position = "none") +
  
  theme(axis.title.y = element_blank())
```

We can clearly see that the Mentally ill, narcotics and alcohol incident rates have been neutral over the year, while gun offenders have recurring peaks in incident rates.

```{r echo=FALSE, message=FALSE, warning=FALSE}
df$date <- substr(df$INCIDENT_DATE, 7, 8)
df_monthdate <-  df %>%
  group_by(date, INC_MONTH) %>%
  summarize(count = n())
df_monthdate <- df_monthdate[complete.cases(df_monthdate), ]
#ggplot(df_monthdate, aes(x= date, y=INC_MONTH,fill = count)) + geom_tile( ) + 
#geom_text(aes(date, INC_MONTH, label = count), color = "black", size = 3) + scale_y_discrete("Months",labels=c("January","February", "March", "April","May", "June","July","August", "September","October","November","December")) + labs(x="Days of Month", y= "Months", title=" Incident Rates across Dates and Months")+
 # scale_fill_gradientn(colours = c("white", "red"))

```
```{r echo=FALSE, message=FALSE, warning=FALSE}
df_monthday <-  df %>%
  group_by(INCIDENT_DAY, INC_MONTH) %>%
  summarize(count = n())

df_monthday <- df_monthday[complete.cases(df_monthday), ]

ggplot(df_monthday, aes(x= INCIDENT_DAY, y=INC_MONTH,fill = count)) + geom_tile( ) + 
geom_text(aes(INCIDENT_DAY, INC_MONTH, label = count), color = "black", size = 4) + scale_y_discrete("Months",labels=c("January","February", "March", "April","May", "June","July","August", "September","October","November","December")) + labs(x="Days of Month", y= "Months", title=" Incident Rates across Dates and Months")+
  scale_fill_gradientn(colours = c("#3794bf", "#FFFFFF", "#df8640"))
```

We can clearly see that the incident rates are more during weekends rather than weekdays from the above graph.

```{r echo=FALSE, message=FALSE, warning=FALSE}
df_subrace <-  df %>%
  group_by(SUBJECT_RACE, INC_MONTH) %>%
  summarize(count = n())

df_offrace <-  df %>%
  group_by(OFFICER_RACE, INC_MONTH) %>%
  summarize(count = n())

df_subrace <- df_subrace[complete.cases(df_subrace), ]
df_offrace <- df_offrace[complete.cases(df_offrace), ]

c1 <- ggplot(df_offrace, aes(x= INC_MONTH, y=OFFICER_RACE,fill = count)) + geom_tile( ) + 
geom_text(aes(INC_MONTH, OFFICER_RACE, label = count), color = "black", size = 4)+xlab("MONTH") +
  scale_fill_gradientn(colours = c("#3794bf", "#FFFFFF", "#df8640"))

c2 <- ggplot(df_subrace, aes(x= INC_MONTH, y=SUBJECT_RACE,fill = count)) + geom_tile( ) + 
geom_text(aes(INC_MONTH, SUBJECT_RACE, label = count), color = "black", size = 4)  +xlab("MONTH")+
  scale_fill_gradientn(colours = c("#3794bf", "#FFFFFF", "#df8640"))

grid.arrange(c1,c2, nrow=2,ncol=1)

```

From the above graph it is very clear that there is a very high chance that a white officer handles the case if the subject is black.

#### 6. Officer experience analysis:

```{r echo=FALSE, message=FALSE, warning=FALSE,results='hide'}
df$OFFICER_HIRE_DATE<-as.Date(df$OFFICER_HIRE_DATE,format="%m/%d/%y")

df$Officer_year <- format(df$OFFICER_HIRE_DATE, "%Y")

df_officer_year <- as.data.frame(df %>% group_by(df$Officer_year) %>% summarise(frequency = n()))

r2 <- ggplot(df_officer_year, aes(x = df_officer_year[,1], y = frequency,group = "count")) + geom_line( size = 1, colour = "orange") +
  geom_point()+theme(axis.text.x = element_text(angle = 90, hjust = 1))+xlab("YEAR")+ylab("OFFICERS HIRED")+ggtitle(" OFFICERS HIRED DURING CORRESPONDING YEARS")
r2
```

We can clearly see a high number of officers hired in the years 2012 to 2015.

```{r echo=FALSE, message=FALSE, warning=FALSE,results='hide'}
#A HISTOGRAM SHOWING OFFICER EXPERIENCE ON FORCE
df$disc_OFFICER_YEARS_ON_FORCE <- as.numeric(df$OFFICER_YEARS_ON_FORCE)
g1<-ggplot(df,aes(df$disc_OFFICER_YEARS_ON_FORCE))+geom_histogram(color="yellow")
g1+xlab("Experience in years")+ylab("No.of officers")+ggtitle(" EXPERIENCE OF OFFICERS")
```

We can see majority number of officers have fewer than 10 years of experience.

```{r echo=FALSE, message=FALSE, warning=FALSE,results='hide'}

arrest=ggplot(df,aes(disc_OFFICER_YEARS_ON_FORCE))+geom_bar()+facet_grid(~SUBJECT_WAS_ARRESTED)+ ylab("Arrested Count")+xlab("Officer Experience")+ ggtitle(" Subjects Arrested vs Officer Experience")
#Bar plot displaying Subject injured Based on the Police Officer Experience
injury=ggplot(df,aes(disc_OFFICER_YEARS_ON_FORCE))+geom_bar()+facet_grid(~SUBJECT_INJURY)+theme(axis.text.x = element_text(angle = 90, hjust = 1)) +xlab("Officer Experience") + ylab("Injured Count")+ ggtitle("Subjects Injured vs Officer Experience")
grid.arrange(arrest,injury)
```

The above figure clearly shows that officers with fewer than 10 years experience have arrested more subjects without any injury on the field.

### Results:

*	The number of incidents has been reducing across the year.
* Crimes reported are less during early mornings and very high during evenings.
*	Crimes were committed more in the month of March.
*	The mean range of incidents is between 3 to 10 for both days and months.
*	Higher crime rates during weekends than weekdays.
*	There is a disparity between the percentages of Black subjects and White officers and Black officers and White subjects.
*	The central division has the highest number of crimes reported.
*	Mentally ill, narcotics and alcohol incident rates have been neutral over the year, while gun offenders have recurring peaks in incident rates.
* It is less likely for an officer with a few years in the force to use violence.

### References:

1.https://moodle.essex.ac.uk/

2.https://www.kaggle.com/
